---
title: "Difference Scores: Paired Samples T-test"
author: "Biostatistics Working Group"
---

# Paired Samples T-test {.unnumbered}

## Overview

<p>

This example examines the question whether participants reported
different mean levels of externalizing problems on the CBCL when
measured at Baseline (T0) compared to when measured at the 1-Year
follow-up (T1). This analysis is conducted in two primary steps: 1)
computing a difference score (DiffScore = CBCL EXT T2 - CBCL EXT T1); 2)
conducting a paired samples t-test on the difference score. Our primary
aim is to determine whether average CBCL externalizing scores of the
participants change from assessement 1 (T1) to assessement 2 (T2), while
accounting for observations that are clustered within youth over time.

</p>

## Preliminary Setup

::: panel-tabset
### Install Packages {.tabset .tabset-fade .tabset-pills}

::: blue
> **This code installs the r packages necessary for this example, if
> they are not already installed**

```{r pckg-install}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

#rm(list = ls())

if (!("lme4" %in% installed.packages())) install.packages("lme4")
if (!("lmerTest" %in% installed.packages())) install.packages("lmerTest")
if (!("tidyverse" %in% installed.packages())) install.packages("tidyverse")
if (!("arrow" %in% installed.packages())) install.packages("arrow")
if (!("arsenal" %in% installed.packages())) install.packages("arsenal")
if (!("kableExtra" %in% installed.packages())) install.packages("kableExtra")
if (!("ggpubr" %in% installed.packages())) install.packages("ggpubr")
if (!("rstatix" %in% installed.packages())) install.packages("rstatix")

```
:::

### Load Packages

::: blue
**This code loads the r libraries necessary for this example**

```{r lib-load}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| class.source: fold-hide

#rm(list = ls())

library(lme4)
library(lmerTest)
library(tidyverse)
library(arrow)
library(arsenal)
library(kableExtra)
library(rstatix)
library(ggpubr)
```
:::

### Config Options

::: blue
**This code configures knitr code chunk options**

```{r config}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
                                  comment=NA, cache=T, code_folding=T,
                                  R.options=list(width=220), fig.align='center', 
                                  out.width='75%', fig.asp=.75)
```
:::
:::

## Descriptives Overview

::: panel-tabset
### Read and View Data {.tabset .tabset-fade .tabset-pills}

::: blue
**This code reads in and shows the data to be used in the current
example**

```{r read-data}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

## Read data
df_long<- read_csv("/Users/shawes/Desktop/data/df_long.csv")
df_wide <- read_csv("/Users/shawes/Desktop/data/df_wide.csv")

# set types
df_long$ids <- as.factor(df_long$ids)
df_long$event <- as.factor(df_long$event)
df_long$site_id <- as.factor(df_long$site_id)
df_long$fam_id <- as.factor(df_long$fam_id)
df_long$sex <- as.factor(df_long$sex)
df_long$age <- as.numeric(df_long$age)
df_long$mature_vg <- as.numeric(df_long$mature_vg)
df_long$cbcl_extern <- as.numeric(df_long$cbcl_extern)
df_long$vg_total <- as.numeric(df_long$vg_total)
```
:::

### Continuous Outcomes

::: blue
**This code creates a table of descriptive information for continuous
outcomes**

```{r contvar-table}
#| messages: FALSE
#| echo: TRUE
#| warning: FALSE
#| results: "asis"

## Create a descriptives table of study variables by measurement occasion 
tab_descriptives_1 <- tableby.control(test=FALSE, total=FALSE, numeric.test="kwt", cat.test="chisq",
numeric.stats=c("N", "meansd", "median", "range"), #"Nmiss2"
cat.stats=c("countpct"), #"Nmiss2"
stats.labels=list(N='Count', meansd="Mean (SD)", median ='Median', 
                  range='Min - Max'))  #, Nmiss2 ='Missing'
my_cont_labels <- list(age = "Age", vg_total = "Weekly # of Video Gaming Hrs",
            cbcl_extern = "CBCL Externalizing Scale"
)

tab_descriptives_1 <- tableby(event ~ age + vg_total + 
                                      cbcl_extern,
                                      data=df_long, control=tab_descriptives_1)

# Push table object through kable and kable_styling
tab_descriptives_1 %>%
  summary(text=TRUE, digits.pct=1, digits=1) %>%
  kable(caption = "Continuous Outcomes") %>%
  kable_styling(bootstrap_options = "striped", 
                full_width = FALSE, html_font = "Cambria",
                font_size = 15, position = "center", fixed_thead = T) %>%
  row_spec(2:3,  bold = F, extra_css = 'vertical-align: middle !important;') %>% 
  column_spec(1, width = "20em", background = "light grey", bold = T, border_right = T) %>%
  column_spec(2, width = "20em", border_right = T) %>%
  column_spec(3, width = "20em", border_right = T) %>%
  footnote(general = "Here is a general comments of the table. ") %>%
  scroll_box(width = "75%", height = "500px")
```
:::

### Categorical Outcomes

::: blue
**This code creates a descriptives table for categorical outcomes**

```{r catvar-table}
#| class.source: 'fold-hide'
#| messages: FALSE
#| warning: FALSE
#| results: "asis"

## Create a descriptives table of study variables by measurement occasion 
tab_descriptives_2  <- tableby.control(test=FALSE, total=FALSE,
numeric.test="kwt", cat.test="chisq",
numeric.stats=c("N", "meansd", "median", "range"
), # "Nmiss2"
cat.stats=c("countpct"), # "Nmiss2"
stats.labels=list(N='Count', meansd="Mean (SD)", median
='Median', range='Min - Max'
)) # , Nmiss2 ='Missing'

my_cat_labels  <- list(
  event = "Year",
  sex = "Sex",
  mature_vg = "Mature Video Games"
  )
          
tab_descriptives_2 <- tableby(event ~ sex + mature_vg, data=df_long, control=tab_descriptives_2)

# Push table object through kable and kable_styling
tab_descriptives_2 %>%
            summary(text=TRUE, digits.pct=1, digits=1) %>%
            kable(caption = "Categorical Outcomes") %>%
            kable_styling(bootstrap_options = "striped", full_width = FALSE, font_size =
                            15, position = "center", fixed_thead = T) %>%
  row_spec(2:3,  bold = F, extra_css = 'vertical-align: middle !important;') %>%
  column_spec(1, width = "20em", background = "light grey", bold = T, border_right = T) %>%
  column_spec(2, width = "20em", border_right = T) %>%
  #column_spec(3, width = "20em", border_right = T) %>%
  footnote(general = "Here is a general comments of the table. ") %>%
  scroll_box(width = "75%", height = "500px")

```
:::
:::

## Results

<div class = "blue">

::: panel-tabset
### Build Model {.tabset .tabset-fade .tabset-pills}

The code snippet below tells R to compute a difference score by
subtracting each participant's Externalizing score at T2 from their
Externalizing score at T1. Next, a one-sample t-test is computed to
examine whether the average difference score is different than zero
(indicating change).

**STEP 1: Compute Difference Score**

```{r build-model-1}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE

# Compute difference score based on CBCL Externalizing subscale scores at baseline (t1) and 1-Year Follow-up (t2)
df_wide$diffscore <- df_wide$cbcl_extern_Year_1 - df_wide$cbcl_extern_Baseline
```

```{r model-checking}
# Compute statistical summaries for the difference score variable
summary(df_wide$diffscore)

```

This summary of the difference score variable indicates xxxxx.

```{r}
hist(df_long$vg_total)

## Summary statistics
summary<-df_long %>%
group_by(event) %>%
get_summary_stats(cbcl_extern, type = "mean_sd")
data.frame(summary)
```

This histogram indicates xxxxxx. The summary command shows xxxxxx.

```{r shapiro-wilks}
### Shapiro-Wilk test and normality (Q-Q) plot (visualization of correlation between a given sample and the normal distribution)
shapiro.test(df_wide$cbcl_extern_Baseline[0:5000])
```

This Shapiro-Wilkes test shows xxxxxx.

```{r qqplot}
qqplot <- ggqqplot(df_wide$cbcl_extern_Baseline,
    ylab = "Externalizing Difference Score", xlab = FALSE,
    ggtheme = theme_minimal()
)

suppressWarnings(print(qqplot)) 
```

This qqplot shows xxxxxx.

**STEP 2: Conduct t-test on Difference Score**

```{r build-model-2}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE
      
# One-sample t-test to determine whether the average difference score is significantly different than 0.
result <- t.test(df_wide$cbcl_extern_Baseline, mu = 0, alternative = "two.sided")
# Print the results
result
```
:::

The output from our model provides: i. a t-statistic ii. degrees of
freedom under iii. p-value (\< .05 then we reject the null hypothesis)
iv. a mean estimate v. 95% confidence intervals

In our example p = .002 which is less than .05 then we reject the null
hypothesis and conclude that the mean difference score Âµ is different
from zero. So we can conclude that thereð· is a difference between cbcl
externalizing scores from t1 to t2. This output also provides the mean
for cbcl externalizing scores (4.45),as well as the correpsonding
confidence intervals

### Model Plots

::: blue
```{r scatterplot}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| code-summary: testing

# Scatterplot to visualize relationship between t1 & t2 data used to create difference score
scatterplot <- ggplot(df_wide, aes(x = cbcl_extern_Baseline, y = cbcl_extern_Year_1)) +
    geom_point(size = 3) +
    geom_smooth(method = lm, se = F) +
    xlab("CBCL Externalizing (x) Baseline") +
    ylab("CBCL Externalizing (x) 1-Year Follow-up")

suppressWarnings(print(scatterplot))
```

Examination of this scatterplot indicates xxxxx.

```{r boxplot}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| code-summary: testing

### boxplot
diffscore_boxplot <- ggboxplot(df_wide$cbcl_extern_Baseline,
    ylab = "Externalizing Difference Score", xlab = FALSE,
    ggtheme = theme_minimal()
)

suppressWarnings(print(diffscore_boxplot))
```

Examination of this boxplot indicates xxxxx.
:::

> This output shows xxxxxxxx. Briefly walk through each metric
:::

## Wrapping Up

<div class = "blue">

::: panel-tabset
### Write-up {.tabset .tabset-fade .tabset-pills}

::: blue
A paired samples t test was conducted to compare differences in
participants cbcl externalzing scores measured across two timepoints.
Findings indicated that there was a difference between cbcl externalzing
scores obtained at T1 compared to cbcl externalzing scores obtained at
T2, t(179) = 3.10, p = .002, Cohen's d = 0.23. Youth exhibited an
increase of cbcl externalzing scores at t2 ( M = 5.67, SD = 1.24)
compared to t1 ( M = 5.83, SD = 1.21).
:::
:::

